---
title: "Coursera Statistical Inference Project \\newline Part 2: Basic inferential data analysis"
author: "Learner42"
date: "19/02/2015"
output: pdf_document
---

# Abstract
In this project, we're going to analyze the ToothGrowth data in the R datasets package. After loading the dataset into R, explanatory analyses are carried out to determine the main trends in the data. In the second parts, statistical methods are applied to confirm or reject these trends.

The source code to this report is hosted on [github](https://github.com/learner42/statinference-011/blob/master/project_2.Rmd).

# Load the data
The data can be loaded via the usual commands.
```{r}
library(datasets)
data(ToothGrowth)
```
Via the help command ``?ToothGrowth`` we know that the dataset contains the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid). 

# Basic exploratory data analyses 
## Group tooth growth by ``method-dose`` pair
As there are 3 dose levels (0.5, 1 and 2) and 2 delivery methods, we can gather basic statistics for these 3 combination. We can split the data per method-dose pair into ``s`` and count the appearances and compute the mean. There are 10 observations per pair.
```{r}
s <- split(ToothGrowth$len, list(ToothGrowth$supp, ToothGrowth$dose))
sapply(s, function(x) { c(count = length(x), mean=mean(x))})
```

## Graphical representation
All data points can be seen via the following graph by
```{r fig.width=8, fig.height=4}
library(ggplot2)
g <- ggplot(data = ToothGrowth, aes(x = dose, y = len, color = supp)) 
g + geom_point() + geom_smooth(method = "lm")
```

## Short summary
Via the grouping and the graphics above, we observe 2 main trends

 * The tooth growth is higher when the dose is higher, providing the method is the same.
 * Delivery via OJ tends to produce faster tooth growth than via VC, with doses at 0.5 and 1.0. With the doses of 2.0 the different between the two method seems to be insignificant.
 
The mean per group obtained by applying ``sapply`` on the splitted data ``s`` seen previouly also confirms this trend.

# Statistical analyses
## Assumptions
In this section, the trends observed in the previous section is evaluated formerly using statistical methods. The analyses are based on the techniques from this class. Regression models, which are also good candidates and are presented in another class of this series, are out of scope of this project.

In order to draw conclusions about the effect of dose and delivery method on tooth growth, let us assume that the tested subects are representative of the species. Since there is no information regarding the order of the data, we consider the data in each of the method-dose pairs independent groups (no way to relate a particular guinea pig among the groups).

## Effect of dose on tooth growth
In this case we only consider tooth growth measurements obtain by the *same method* (OJ or VC). With each method, we compare the 2 pairs: tooth growth of ``dose 0.5`` vs. ``dose 1``; and ``dose 1`` vs. ``dose 2``. There are then 4 tests in total. 

For example, let us consider the **VC method** with dose 0.5 and 1:

 * Null hypothesis $\mu_0$: using a dose of 0.5 or 1.0 produces the same tooth growth.
 * Alternative hypothesis $\mu_a$: A dose of 1.0 produce higher tooth growth than a dose of 0.5.

Let us remind ourselves that the splitted data are stored in ``s``. We can than apply the ``t.test`` function to obtain the result

```{r}
t.test(s$VC.1, s$VC.0.5, paired = F, alt = "greater")$p.value
```

The result is way below the limit ($\alpha = 0.05$ for one sided test et 95% confidence), we can then reject the null hypothesis. In other words, with "Vitamin C" delivery method, a dose of 1.0 mg produces higher tooth growth than a dose of 0.5 mg.

### Summary
```{r echo=FALSE}
get.p.value <- function(lab1, lab2) {
  t.test(s[[lab1]], s[[lab2]], paired = F, alt = "greater")$p.value
}
```  

| Group 1    | Group 2     | p-value                              | t-test outcome   | Conclusion
|------------|-------------|--------------------------------------|------------------|---------------
| VD - 1 mg  | VD - 0.5 mg | $`r get.p.value("VC.1",  "VC.0.5")`$ | Reject null hyp. | Group 1 produces higher tooth growth
| VD - 2 mg  | VD - 1   mg | $`r get.p.value("VC.2",  "VC.1"  )`$ | Reject null hyp. | Group 1 produces higher tooth growth
| OJ - 1 mg  | OJ - 0.5 mg | $`r get.p.value("OJ.1",  "OJ.0.5")`$ | Reject null hyp. | Group 1 produces higher tooth growth
| OJ - 2 mg  | OJ - 1   mg | $`r get.p.value("OJ.2",  "OJ.1"  )`$ | Reject null hyp. | Group 1 produces higher tooth growth

The hypotheses are the same as above, namely: null hypothesis: both groups produce the same effect; alternative hypothesis: Group 1 produces higher tooth growth. In all cases, we reject the null hypothesis: among the tested scenarios, **higher dose produces higher tooth growth**. 

## Effect of delivery method on tooth growth
In this case we only consider tooth growth measurements obtain with the *same dose*. With each dose, we compare the VC and OJ method: there are 3 tests in total. Same reasoning as the previous section.

### Summary
| Group 1     | Group 2     | p-value                              | t-test outcome   | Conclusion
|-------------|-------------|--------------------------------------|------------------|---------------
| OJ - 0.5 mg | VD - 0.5 mg | $`r get.p.value("OJ.0.5",  "VC.0.5")`$ | Reject null hyp. | Group 1 produces higher tooth growth
| OJ - 1   mg | VD - 1   mg | $`r get.p.value("OJ.1",    "VC.1"  )`$ | Reject null hyp. | Group 1 produces higher tooth growth
| OJ - 2   mg | VD - 2   mg | $`r get.p.value("OJ.2",    "VC.2")`$ | Accept null hyp. | Both groups produces the same effect

At smaller doses, i.e. 0.5 and 1 mg, "orange juice" method produces higher tooth growth than the "vitamin C" method. At higher dose (2 mg) both methods are equally effective.
